Data-driven approaches to identifying knowledge gaps in marine invertebrate biodiversity

### 2017-05-26 manuscript.Rmd
edit manuscript

# Main text

@@

@@

Until very recently, our knowledge of biodiversity was scattered among heterogeneous sources (e.g. natural history collections, publications, expert knowledge), and difficult to access. These factors have contributed to the lack of synthesis and have made the identification of knowledge gaps difficult. However, with the development of public databases for natural history records (e.g. iDigBio [@Page2015]), species occurrences (e.g., GBIF [@Edwards2004], OBIS [@Zhang2002]), genetic data (e.g., GenBank [@Benson2013], BOLD [@Ratnasingham2007a]), and registers of species names (e.g., WoRMS [@Costello2013b]), we can quantitavely evaluate what we know. Here, we aggregate publicly available data for marine invertebrates that occur in the United States and in the Gulf of Mexico to highlight gaps in our knowledge. We show that public databases are relatively incomplete, and that most species are known from a few speciemens and a few locations. We emphasize that our understanding of marine biodiversity, and the impact of human activities on communities require not only to get a more accurate list of the species that inhabits the oceans, but we also need more comprehensive data on species distributions and abundances. This can only be achieved with additional sampling and a commitment to training more taxonomists and experts that can identify the material collected.

Regional faunistic lists assembled and curated by taxonomic experts typically provide a comprehensive and detailed account of species that are known to occur in a given geographical area. We assessed how faunistic lists assembled over decades by specialists compare with records in public databases. We parsed faunistic lists for marine invertebrates from the Gulf of Mexico [@Tunnell2009] and from the Pacific Northwest [@Kozloff1987]. Our goals were (1) to compare  the species lists for these geographical areas assembled by experts with the records found in databases; (2) to assess the proportion of species included in the expert lists that have records in public databases.

To compare the faunistic lists from the Gulf of Mexico and the Pacific Northwest with records in public databases, we queried the iDigBio and OBIS databases for all records they included in the same geographic boundaries as the lists. After filtering these lists to only include the same higher taxonomic level taxa as the species lists, we had for the Gulf of Mexico `r nrow(idigbio_gom_records)` and `r nrow(obis_gom_records)` records for iDigBio and OBIS respectively representing `r dplyr::n_distinct(idigbio_gom_records, worms_valid_name)` and `r dplyr::n_distinct(obis_gom_records, worms_valid_name)` species. For the Gulf of Mexico, most of the species included in the expert lists are recorded in both iDigBio and OBIS (xx %), or OBIS (xx %). However, for the Pacific Northwest, only xx% of the species from the expert list have records in public databases.

assess the proportion of species included in these lists that (1) are recorded in databases from the same geographical area; (2) have any records in databases
, with the records that occur in the same geographical areas in iDigBio and OBIS. Our goal was to assess the proportion of species known to occur in a given area by specialists that have been captured by digitization efforts. First, to estimate the proportion of species included in these faunistic lists whose occurences have been digitized, we queried biodiversity databases using the same geographical limits as the ones used by the authors of these faunistic lists. We found ... Second, to estimate whether listed species are known to at least occur in these databases, we queried them for the presence of the species within the databases.

We extracted `r sum(gom_worms$is_marine, na.rm = TRUE)` marine invertebrates species from the faunistic list of the Gulf of Mexico, and `r sum(kozloff_worms$is_marine, na.rm=TRUE)` species from the Salish Sea. Species are unevenly represented in databases, with 25% to 50% of species listed to occur in these locations, that are not represented in digital records of museum collections. Typically a higher proportion of species from the Pacific Coast than the Gulf of Mexico are not represented. Comparatively, all but 2 of the `r nrow(mammals_species_asm)` mammal species and `r sum(amphibian_summary$count == 0)` of the `r nrow(amphibian_summary)` amphibian species known to occur in the US have digitized museum specimens.

The proportion of species that have DNA barcodes available is surprisingly constant across the better studied phyla. About 20% of species known to occur in the Gulf of Mexico and in the Salish Sea have DNA barcode records.  Because geographic information is missing for most records (`r format_output(100*stat_bold$p_no_coords)`% of the `r format(stat_bold$n_total_records)` records analyzed here have no geographical coordinates information associated with them), and many marine invertebrate species are known to be cryptic complexes that are not as widely distributed as previously thought. Therefore specimens identified under the same species but sampled in distant locations, could represent cryptic species. It is thus crucial to ensure that accurate geographic information is provided when DNA barcodes are published, and that additional samples are collected, to capture more accurately their geographic range and the extent of their genetic diversity. This high level of cryptic diversity is reflected in the fact that `r stat_bold$n_spp_muti_bin` species out of `r stat_bold$n_spp` include more than 1 BIN. This number is confounded by some taxonomic identification issues as `r stat_bold$n_shared_bins` BINs are shared across more than one species.

<!-- However, among these records, xx% have been sampled in the US (xx%), Mexico (xx%) or Canada (xx%).  -->
<!-- Paragraph about the same numbers but based on species list from iDigBio: -->

<!-- - how many species are in iDigBio and don't show up in species lists from Kozloff and GOM -->
<!-- - how many species in iDigBio have barcodes -->
<!-- - how many species in iDigBio have OBIS and GBIF records -->

@@

@@

However, taxonomic expertise is dwindling, and there are now only a handful of US natural history museums with active marine invertebrates collections. Museum collections have been recording our knowledge about species and their geographical ranges, but until recently, this information was fragmented across institutions and difficult to synthesize. With collection digitization and other large-scale projects have aimed at gathering biodiversity information, public databases can now be used to enable synthesis and highlight knowledge gaps. Here we use these data to synthetic the state of knowledge for marine invertebrates, and demonstrate that the stage is set to combine these data sources to inform where future efforts for documenting biodiversity.

@@

>>>>

@@

## Abstract

@@

Until very recently, our knowledge of biodiversity was scattered among heterogeneous sources (e.g. natural history collections, publications, expert knowledge), and difficult to access. These factors have contributed to the lack of synthesis and have made the identification of knowledge gaps difficult. However, with the development of public databases for natural history records (e.g. iDigBio [@Page2015]), species occurrences (e.g., GBIF [@Edwards2004], OBIS [@Zhang2002]), genetic data (e.g., GenBank [@Benson2013], BOLD [@Ratnasingham2007a]), and registers of species names (e.g., WoRMS [@Costello2013b]), we can now evaluate what we know. Here, we aggregate publicly available data for marine invertebrates that occur in the United States, in the Gulf of Mexico, and in the Pacific North-West, to highlight gaps in our knowledge. We show that public databases are relatively incomplete, and that most species are known from a few speciemens and a few locations. We emphasize that our understanding of marine biodiversity, and the impact of human activities on communities require not only to get a more accurate list of the species that inhabits the oceans, but we also need more comprehensive data on species distributions and abundances. This can only be achieved with additional sampling and a commitment to training more taxonomists and experts that can identify the material collected.

## Results

### Comparison with Regional Lists by Geography

Regional faunistic lists assembled and curated by taxonomic experts typically provide a comprehensive and detailed account of species that are known to occur in a given geographic area. We assessed how faunistic lists assembled over decades by specialists compare with records in public databases for the same geographic area. We parsed faunistic lists to extract marine invertebrates from the Gulf of Mexico [@Tunnell2009] and from the Pacific Northwest [@Kozloff1987]. Our goals were (1) to compare  the species lists for these geographic areas assembled by experts with the records found in databases; (2) to assess the proportion of species included in the expert lists that have records in public databases.

To compare the faunistic lists from the Gulf of Mexico and the Pacific Northwest with records in public databases, we queried the iDigBio and OBIS databases for all records they included in the same geographic boundaries as the lists. After filtering these lists to only include the same higher taxonomic level taxa as the species lists, we had for the Gulf of Mexico `r nrow(idigbio_gom_records)` and `r nrow(obis_gom_records)` records for iDigBio and OBIS respectively representing `r dplyr::n_distinct(idigbio_gom_records$worms_valid_name)` and `r dplyr::n_distinct(obis_gom_records$worms_valid_name)` species. For the Gulf of Mexico, most of the species included in the expert lists are recorded in both iDigBio and OBIS (xx %), or OBIS (xx %). However, for the Pacific Northwest, only xx% of the species from the expert list have records in public databases (Fig. 1).

Species that are included in the list but do not have records in databases could be (1) rare: they have never been collected or have been collected only a few times, and the corresponding records have not been digitized yet; (2) difficult to preserve: large, very small, or fragile species are preserved less frequently; (3) are in the database under a different name: the species has been synonimized or re-identified since the publication of the list under a different name that may be included in the database; (4) at the limit of its range: it was included in the list because it is known to occur at the limit of its geographic zone, but has never been collected within the boundaries.

Species that are in the database but not in the list could be: (1) records not considered by the authors of the checklist: either because they were not accessible or known to the authors; (2) records erronously excluded the records (2) records that have been discovered since the list has been assembled; (3) records with incorrect identification; (4) records with incorrect geographic coordinates: their associated geographic information in the database places them in the Gulf of Mexico or the Pacific Northwest, but they were actually collected elsewhere.

### Comparison with Regional Lists by Species Names

To assess whether species included in regional lists but absent from the same geographic area nevertheless occur elsewhere, we queried the iDigBio and OBIS database for all the marine invertebrate species included in these lists.

We extracted `r sum(gom_worms$is_marine, na.rm = TRUE)` marine invertebrates species from the faunistic list of the Gulf of Mexico, and `r sum(kozloff_worms$is_marine, na.rm=TRUE)` species from the Salish Sea. We then looked for the number of species listed that are known to occur in the same geographic area, the United States, or anywhere else in the world. Overall, OBIS had records for most of the species that are included in the species lists. However, even for the better studied phyla about 75% of the species have records from the same geographic area, while it is less than 50% for the Pacific Northwest.

These results suggest that even when considering species lists that have been curated and that represent possibly the most accurate knowledge for this geographic area, it would not be possible to reconstruct these lists based on currently digitized information. It highlights that additional digitization but even more importantly additional collecting is needed to validate the occurences of the species included in these lists.

Species are unevenly represented in databases, with 25% to 50% of species listed to occur in these locations, that are not represented in digital records of museum collections. Typically a higher proportion of species from the Pacific Coast than the Gulf of Mexico are not represented. Comparatively, all but 2 of the `r nrow(mammals_species_asm)` mammal species and `r sum(amphibian_summary$count == 0)` of the `r nrow(amphibian_summary)` amphibian species known to occur in the US have digitized museum specimens.

### What do we know about marine invertebrates in the US?

@@

Based on these data, there is no evidence of a diversity gradient.

@@

However, taxonomic expertise is dwindling, and there are now only a handful of US natural history museums with active marine invertebrates collections. Museum collections have been recording our knowledge about species and their geographic ranges, but until recently, this information was fragmented across institutions and difficult to synthesize. With collection digitization and other large-scale projects have aimed at gathering biodiversity information, public databases can now be used to enable synthesis and highlight knowledge gaps. Here we use these data to synthetic the state of knowledge for marine invertebrates, and demonstrate that the stage is set to combine these data sources to inform where future efforts for documenting biodiversity.

### Current genetic data and the future of biodiveristy discovery and monitoring

The proportion of species that have DNA barcodes available is surprisingly constant across the better studied phyla. About 20% of species known to occur in the Gulf of Mexico and in the Salish Sea have DNA barcode records.  Because geographic information is missing for most records (`r format_output(100*stat_bold$p_no_coords)`% of the `r format(stat_bold$n_total_records)` records analyzed here have no geographic coordinates information associated with them), and many marine invertebrate species are known to be cryptic complexes that are not as widely distributed as previously thought. Therefore specimens identified under the same species but sampled in distant locations, could represent cryptic species. It is thus crucial to ensure that accurate geographic information is provided when DNA barcodes are published, and that additional samples are collected, to capture more accurately their geographic range and the extent of their genetic diversity. This high level of cryptic diversity is reflected in the fact that `r stat_bold$n_spp_muti_bin` species out of `r stat_bold$n_spp` include more than 1 BIN. This number is confounded by some taxonomic identification issues as `r stat_bold$n_shared_bins` BINs are shared across more than one species.

## What to do next?

* Improve data quality:
  - easy: fix mispelling in identifications
  - improve identification of unidentified lots (requires taxonomists/specialists to be able to identify the specimens)
  - add and validate geographic information associated with specimens
  - develop infrastructure and tools to continuously check and validate data
    quality in databases.

@@

## Conclusions
